\emph{icepack}: a novel glacier flow modeling package

### 2019-03-22 icepack.tex
Started writing about model substitution

There are two diagnostic models implemented in icepack.
The shallow stream approximation (SSA) is a 2D model describing the depth-averaged velocity of a fast-flowing grounded ice stream or floating ice shelf \citep{macayeal1989large}.
The SSA model is appropriate where the sliding velocity is close to the surface velocity, or in other words where the ice is nearly in plug flow.
Plug flow is a good approximation near the margins of an ice sheet, but deep in the interior the flow is mostly by vertical shear.
The first-order or Blatter-Pattyn (BP) approximation, by contrast, is a 3D model describing the horizontal velocity \citep{blatter1995velocity, pattyn2003new}.
The only approximation in the BP model is that the flow has a low aspect ratio -- the thickness of the glacier is much less than its horizontal extent.
This approximation may be questionable around, say, the main trunk of Jakobshavn Isbrae in Greenland, which flows through a very deep and narrow trough.
Even Jakobshavn has an aspect ratio on the order of 1/5 and almost all glacier flows have an aspect ratio less than 1/10 or even 1/20.

@@

\subsection{Substitution of model components}

>>>>

There are two diagnostic models implemented in icepack.
The \emph{shallow stream approximation} (SSA) is a 2D model describing the depth-averaged velocity of a fast-flowing grounded ice stream or floating ice shelf \citep{macayeal1989large}.
The SSA model is appropriate where the sliding velocity is close to the surface velocity, or in other words where the ice is nearly in plug flow.
Plug flow is a good approximation in fast-flowing ice streams and outlet glaciers near the margins of an ice sheet, but deep in the interior the flow is mostly by vertical shear.
The \emph{first-order} or \emph{Blatter-Pattyn} (BP) approximation, by contrast, is a 3D model describing the horizontal velocity \citep{blatter1995velocity, pattyn2003new}.
The only approximation in the BP model is that the flow has a low aspect ratio -- the thickness of the glacier is much less than its horizontal extent.
This approximation may be questionable around, say, the main trunk of Jakobshavn Isbrae in Greenland, which flows through a very deep and narrow trough.
Even Jakobshavn has an aspect ratio on the order of 1/5 and almost all glacier flows have an aspect ratio less than 1/10 or even 1/20.

@@

\subsection{Substituting model components}

Many aspects of glacier physics are not completely understood.
For example, the most common description of basal friction assumes that glacier sliding occurs through regelation, in which case the basal shear stress can be written
\begin{equation}
    \tau_b = -C|u|^{\frac{1}{m} - 1}u
\end{equation}
with $m = 3$ \citep{weertman1957sliding}.
Many authors have argued that, in fast-flowing regions of the ice sheet, glacier sliding occurs instead by plastic deformation within the subglacial sediments \citep{tulaczyk2000basal}.
For plastic sliding, the shear stress is dependent on the yield strength of the subglacial sediments and not on the sliding speed, in which case $m = \infty$ would be most appropriate.
Finally, the Schoof sliding law aims to reconcile the two modes of sliding \citep{schoof2005effect}.
Below a critical sliding speed, the shear stress is aymptotic to $|u|^{1/m}$ as in the Weertman sliding law, but above the critical speed the shear stress is independent of the sliding speed.

The Weertman and plastic sliding laws possess the same functional form but differ only in the value of a single scalar parameter $m$.
The Schoof sliding law, on the other hand, has a totally different functional dependence on the velocity.
Several authors, including Schoof, have proposed that the basal shear stress is also a function of the effective pressure $N = \rho gh - p_w$ within the subglacial hydrological system \citep{budd1979empirical, schoof2005effect}.
Implementing these more sophisticated mathematical models would require adding an extra argument to the procedure for solving the diagnostic equation.

One of our goals with icepack is to facilitate experimentation with the model physics.
To support use cases like implementing the Schoof sliding law, it must be possible to completely alter the functional form of a given model physics component.
For uses cases like explicitly adding the dependence of basal shear stress on hydrology, it must also be possible to add entirely new fields to a given model physics component.
In programming terms this amounts to changing the number of arguments to the function that calculates basal shear stress.
For a library developed in C or Fortran, the user would then also have to change the signature of the diagnostic solve function.
In C++ one could obviate the need to modify the diagnostic solve routine by using variadic templates.
But variadic templates are a more advanced language feature and using them would arguably present a steep learning curve for novices.

In icepack, any given model physics component can be substituted for a parameterization of the user's choosing.
Every flow model in icepack is represented by a Python object, and each model object is initialized with a set of functions to calculate the various terms of the action functional.
For example, the ice stream model is initialized with functions to calculate the contributions due to viscosity, basal friction, side wall drag, and gravity.
Sensible parameterizations are selected by default.
If the user does wish to change a given component, they can pass a function that they write to the initializer for the model object.
When the diagnostic solve procedure is invoked, all of its arguments are then passed on by keyword to the method that calculates the action functional.
Any fields that are unnecessary for calculating a given component are simply ignored.
For example, the rheology and friction coefficient are passed to the function that calculates the gravitational driving power, and this function simply makes no use of these arguments.
The functional form can be changed entirely and new arguments can be added, with the only restriction that they are passed as keyword and not as positional arguments.

This approach to making the physics models more extensible restricts us to using keyword arguments instead of positional arguments.
In this case, passing arguments by keyword enhances readability and comprehensibility.
The user only needs to know the argument names, which are chosen to agree with what symbols are commonly used for each ($C$ for friction, $A$ for rheology, $N$ for effective pressure, etc.).
The order of the arguments is arbitrary and immaterial.
For defining, say, the bilinear form that represents a non-symmetric PDE, the order of the arguments does have some inherent mathematical significance.
